Light and life form interact to shape stomatal ratio among British angiosperms$^1$

### 2017-07-13 ms/ms.Rnw
edited introduction

@@

% OPTIONAL
% Synthesis model
% total stomatal density versus growth form (as supplement???)
% used ordered multinomial rather than lm for lf-light corr?
% COMPADRE?
% Model adequacy?
% FIGURE FOR SEM?

@@

@@

\section*{Acknowledgements}

% \section*{Author contribution statement}
% CDM designed the study, collected data, analyzed the data, and wrote the manuscript.

@@

\begin{itemize}
	\item In most plants, stomata are located only on the abaxial leaf surface (hypostomy), but many plants have stomata on both surfaces (amphistomy). Variation in stomatal ratio (the ratio of ab- and adaxial stomatal densities) is probably adaptive, but the ecological conditions that favor amphistomy are not well understood. In particular, high light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together.
	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, Raunki\ae r lifeform, and phylogenetic relationships on \Sexpr{nrow(stomata)} species of British angiosperms. I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio.
	\item (return to this) key results: \el, growth form, and interaction are important
	\item I show for the first time that light and growth form interact to shape variation in stomatal ratio; amphistomy is advantageous in high light, but mostly for herbs. These results improve our understanding of the adaptive significance of stomatal ratio, use stomatal ratio as proxy for paleo vegetation, and as a target for crop improvement.
\end{itemize}

@@

While the density and size of stomata have been researched extensively [CITATIONS], the adaptive significance of stomatal distribution is less well understood. Stomata are most often found only on the lower leaf surface (hypostomy), but occur on both surfaces (amphistomy) in many species \citep{Metcalfe_Chalk_1950, Parkhurst_1978, Mott_etal_1982}. Theory and experiments demonstrate that amphistomy increases photosynthetic rates under many conditions. By creating a second parallel pathway for CO$_2$ diffusion within the mesophyll, amphistomy optimally supplies CO$_2$ \citep{Parkhurst_1978, Gutschick_1984b, Jones_1985}. Amphistomy is correlated with greater CO$_2$ diffusion \citep{Beerling_Kelly_1996} and higher photosynthetic rates \citep{Mckown_etal_2014}. These observations are corroborated by experiments demonstrating that amphistomy increases maximum photosynthetic rates by up to 20\% \citep{Parkhurst_Mott_1990}. However, amphistomy can increase transpiration \citep{Jones_1985, Foster_Smith_1986}, though empirical studies suggest great water-use efficiency in amphistomatous species \citep{Bucher_etal_2017}. Hence, amphistomy clearly should be favored when CO$_2$ limits photosynthetic rate, but the open questions are under what ecological conditions does CO$_2$ supply most strongly limit photosynthetic rate \citep{Peat_Fitter_1994b} and when is photosynthetic rate most important to fitness?

@@

%%%%% PARAGRAPH 3-4: adaptive hypotheses and current limitations %%%%%
The leading, nonmutually exclusive hypotheses are that 1) open habitats favour amphistomy because CO$_2$ diffusion most strongly limits photosynthetic rate under high light and 2) herbaceous growth form favours amphistomy because traits that maximize photosynthetic rate are often under stronger selection in herbs. \citeauthor{Salisbury_1927} (\citeyear{Salisbury_1927}) first noted that amphistomy is most common in herbaceous plants from open habitat (i.e. high light) of the British flora. These observations have been replicated in other studies \citep{Mott_etal_1982, Peat_Fitter_1994b, Jordan_etal_2014, Muir_2015} and may support physiological and ecological hypotheses that CO$_2$ most strongly limits photosynthesis in high light and/or photosynthesis contributes most to fitness in herbaceous plants. Under high light, CO$_2$ can strongly limit maximum photosynthetic rates, espcecially in thick leaves \citep{Jones_1985}. Hence, having stomata on both surfaces relieves this limitation by adding a second parallel pathway for CO$_2$ diffusion. Parkhurst \citeyear{Parkhurst_1978} argued that greater leaf thickness \textit{per se} selected for amphistomy, but there is little evidence for correlations between leaf thickness and stomatal ratio independent of light \citep{Mott_etal_1982, Gibson_1996, Muir_2015}. Amphistomy is correlated with open habitat in warm desert plants of western North America \citep{Mott_etal_1982, Gibson_1996}, among the Proteaceae \citep{Jordan_etal_2014}, and in continental European herbs \citep{Bucher_etal_2017}.

Stomatal ratio is also associated with growth form. In the British flora, \citeauthor{Salisbury_1927} (\citeyear{Salisbury_1927}) found that trees and shurbs are nearly always hypostomatous, whereas herbs from open habitat are amphistomatous. This pattern holds when data are averaged by family to coarsely control for phylogenetic nonindependence \citep{Peat_Fitter_1994b} or when using alternative classification schemes, such as Raunki\ae r life form \citep{Peat_Fitter_1994b}. Across plants from ~90 familes worldwide, growth form is the strongest predictor of stomatal ratio when multiple factors are estimated simultaneously and controlling for phylogenetic nonindependence \citep{Muir_2015}. These patterns are consistent with other data indicating that many herbaceous plants are under strong selection for high maximum photosynthetic rates. \citep{Bazzaz_1979, Korner_etal_1989}. NEED MORE RECENT CITATIONS ON THIS.

@@

%%%% PARAGRAPH 5-6 %%%%%
Although previous comparative studies have tested whether open habitat and growth form influence stomatal ratio, we do not know if these effects are independent of one another. Open habitat and growth form are probably not independent because open habitat by definition consists of more short-statured, herbaceous plants. Even attempts to disentangle light and growth form by contrasting herbs from open and understory habiats \citep{Salisbury_1927} are problematic because high light habitats select for faster life history stratagies \citep{Galloway_Etterson_2007} [<-- OTHER REFS IN THAT PAPER?]. Open habitat and groth form may also interact with one another. For example, amphistomy may only favored when CO$_2$ strongly limits photosynthetic rate \textit{and} photosynthetic rate strongly limits fitness.

@@

To address these questions, I reanalyzed existing data on stomatal ratio, light habitat, and growth form in British angiosperms \citep{Fitter_Peat_1994a, BEF} using phylogenetic comparative methods. The British angiosperm flora is well suited for these questions because this flora has been comprehensively surveyed for many ecologically important traits, meaning it is probably the least biased survey of stomatal trait variation. Salisbury's observations on stomata and ecology in the British flora have heavily influenced plant ecophysiology, but many of his and subsequent authors' analyses have significant limitations because of inadequate statistical methods. For example, few analyses until recently account for phylogenetic nonindependence \citep{Felsenstein_1985}, which can strongly influence inferences on stomatal traits and growth form \citep{Kelly_Beerling_1995}. A species-level phylogeny of the entire British flora \citep{Lim_etal_2014} now allows me to rigorously analyze evolutionary relationships between stomatal ratio, light, and growth form.

@@

>>>>

% Things to do before submission:
% Sally's notes
% Matt's notes
% Remove code to copy refs
% Github DOI?
% Finish abstract
% Finish citations
% Licenses (see http://ropensci.github.io/reproducibility-guide/sections/introduction/)

@@

@@

  # Remove before publication
  file.copy(from = "~/Google Drive/CommonResources/refs.bib",
            to = "~/Google Drive/britstom/ms/refs.bib", overwrite = TRUE)

@@

\section*{Acknowledgements}
I thank Sally Otto and Matt Pennell for feedback on this manuscript.
% \section*{Author contribution statement}
% CDM designed the study, collected data, analyzed the data, and wrote the manuscript.

@@

\begin{itemize}
	\item In most plants, stomata are located only on the abaxial leaf surface (hypostomy), but many plants have stomata on both surfaces (amphistomy). Variation in stomatal ratio (the ratio of ab- and adaxial stomatal densities) is probably adaptive, but the ecological conditions that favor amphistomy are not well understood. In particular, high light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together.
	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, Raunki\ae r lifeform, and phylogenetic relationships for \Sexpr{nrow(stomata)} species of British angiosperms. I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio.
	\item (return to this) key results: \el, growth form, and interaction are important
	\item I show for the first time that light and growth form interact to shape variation in stomatal ratio; amphistomy is advantageous in high light, but mostly for herbs. These results improve our understanding of the adaptive significance of stomatal ratio, use stomatal ratio as proxy for paleo vegetation, and as a target for crop improvement.
\end{itemize}

@@

While the density and size of stomata have been researched extensively [CITATIONS], the adaptive significance of stomatal distribution is less well understood. Stomata are most often found only on the lower leaf surface (hypostomy) but occur on both surfaces (amphistomy) in many species \citep{Metcalfe_Chalk_1950, Parkhurst_1978, Mott_etal_1982}. Theory and experiments demonstrate that amphistomy increases photosynthetic rates under many conditions. By creating a second parallel pathway for CO$_2$ diffusion within the mesophyll, amphistomy optimally supplies CO$_2$ \citep{Parkhurst_1978, Gutschick_1984b, Jones_1985}. Amphistomy is correlated with greater CO$_2$ diffusion \citep{Beerling_Kelly_1996} and higher photosynthetic rates \citep{Mckown_etal_2014}. These observations are corroborated by experiments demonstrating that amphistomy increases maximum photosynthetic rates by up to 20\% \citep{Parkhurst_Mott_1990}. On the other hand, amphistomy can increase transpiration \citep{Jones_1985, Foster_Smith_1986, Buckley_etal_2015}. While transition to amphistomy is thus thought to increase transpiration, empirical studies suggest greater water-use efficiency in amphistomatous species \citep{Bucher_etal_2017}. Hence, amphistomy appears to benefit a plant's carbon use relative to water loss and should be favored when CO$_2$ limits photosynthetic rate. The open questions are under what ecological conditions does CO$_2$ supply most strongly limit photosynthetic rate \citep{Peat_Fitter_1994b} and when is photosynthetic rate most important to fitness?

@@

%%%%% PARAGRAPH 3-4: adaptive hypotheses and current limitations %%%%%
The leading, nonmutually exclusive hypotheses are that 1) open habitats favour amphistomy because CO$_2$ diffusion most strongly limits photosynthetic rate under high light and 2) herbaceous growth form favours amphistomy because traits that maximize photosynthetic rate are often under stronger selection in herbs. \citeauthor{Salisbury_1927} (\citeyear{Salisbury_1927}) first noted that amphistomy is most common in herbaceous plants from open habitats (i.e., with high light) of the British flora. These observations have been replicated in other studies \citep{Mott_etal_1982, Peat_Fitter_1994b, Jordan_etal_2014, Muir_2015} and may support physiological and ecological hypotheses that CO$_2$ most strongly limits photosynthesis in high light and/or photosynthesis contributes most to fitness in herbaceous plants. Under high light, CO$_2$ can strongly limit maximum photosynthetic rates, espcecially in thick leaves \citep{Jones_1985}. Hence, having stomata on both surfaces relieves this limitation by adding a second parallel pathway for CO$_2$ diffusion. Parkhurst \citeyear{Parkhurst_1978} argued that greater leaf thickness \textit{per se} selected for amphistomy, but there is little evidence for correlations between leaf thickness and stomatal ratio independent of light \citep{Mott_etal_1982, Gibson_1996, Muir_2015}. Amphistomy is correlated with open habitat in warm desert plants of western North America \citep{Mott_etal_1982, Gibson_1996}, among the Proteaceae \citep{Jordan_etal_2014}, and in continental European herbs \citep{Bucher_etal_2017}.

Stomatal ratio is also associated with growth form. In the British flora, \citeauthor{Salisbury_1927} (\citeyear{Salisbury_1927}) found that trees and shurbs are nearly always hypostomatous, whereas herbs from open habitats are amphistomatous. This pattern holds when data are averaged by family to coarsely control for phylogenetic nonindependence \citep{Peat_Fitter_1994b} or when using alternative classification schemes, such as Raunki\ae r life form \citep{Peat_Fitter_1994b}. Across plants from ~90 familes worldwide, growth form is the strongest predictor of stomatal ratio when multiple factors are estimated simultaneously and controlling for phylogenetic nonindependence \citep{Muir_2015}. These patterns are consistent with other data indicating that many herbaceous plants are under strong selection for high maximum photosynthetic rates. \citep{Bazzaz_1979, Korner_etal_1989}. NEED MORE RECENT CITATIONS ON THIS.

@@

%%%% PARAGRAPH 5-6 %%%%%
Although previous comparative studies have tested whether open habitat and growth form influence stomatal ratio, we do not know if these effects are independent of one another. Open habitat and growth form may not be independent because open habitats generally consist of more short-statured, herbaceous plants. Some authors have attempted to disentangle light and growth form by contrasting herbs from open and understory habiats \citep{Salisbury_1927}. However, this is problematic if phylogenetic relationships are not controlled for, because shade species may share traits simply because they are more closely related to each other than they are to high light species. Finally, open habitat and groth form may also interact with one another. For example, amphistomy may only be favored when CO$_2$ strongly limits photosynthetic rate (e.g. in high light) \textit{and} photosynthetic rate strongly limits fitness (e.g. in herbs).

@@

To address these questions, I reanalyzed existing data on stomatal ratio, light habitat, and growth form in British angiosperms \citep{Salisbury_1927, Fitter_Peat_1994a, BEF} using phylogenetic comparative methods. The British angiosperm flora is well suited for these questions because this flora has been comprehensively surveyed for many ecologically important traits, meaning it is probably the least biased survey of stomatal trait variation. Salisbury's observations on stomata and ecology in the British flora have heavily influenced plant ecophysiology, but many of his and subsequent authors' analyses have significant limitations because of inadequate statistical methods. For example, few analyses until recently account for phylogenetic nonindependence \citep{Felsenstein_1985}, which can strongly influence inferences on stomatal traits and growth form \citep[this study did not consider light]{Kelly_Beerling_1995}. A species-level phylogeny of the entire British flora \citep{Lim_etal_2014} now allows for the first time rigorous analysis of evolutionary relationships among stomatal ratio, light, and growth form.

% NOTES , for example, two herbaceous species adapted to different light conditions may not really be comparable since light has a strong impact on life history selection \citep{Franco_Silvertown_1996, Galloway_Etterson_2007}.

@@

Data and annotated source code to generate this manuscript are available on GitHub (https://github.com/cdmuir/britstom) and Dryad \citep{Muir_dryad}.

### 2017-07-13 ms/ms.tex
edited introduction

@@

\begin{itemize}
	\item In most plants, stomata are located only on the abaxial leaf surface (hypostomy), but many plants have stomata on both surfaces (amphistomy). Variation in stomatal ratio (the ratio of ab- and adaxial stomatal densities) is probably adaptive, but the ecological conditions that favor amphistomy are not well understood. In particular, high light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together.
	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, Raunki\ae r lifeform, and phylogenetic relationships on 372 species of British angiosperms. I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio.
	\item (return to this) key results: \el, growth form, and interaction are important
	\item I show for the first time that light and growth form interact to shape variation in stomatal ratio; amphistomy is advantageous in high light, but mostly for herbs. These results improve our understanding of the adaptive significance of stomatal ratio, use stomatal ratio as proxy for paleo vegetation, and as a target for crop improvement.
\end{itemize}

@@

To address these questions, I reanalyzed existing data on stomatal ratio, light habitat, and growth form in British angiosperms \citep{Fitter_Peat_1994a, BEF} using phylogenetic comparative methods. The British angiosperm flora is well suited for these questions because this flora has been comprehensively surveyed for many ecologically important traits, meaning it is probably the least biased survey of stomatal trait variation. Salisbury's observations on stomata and ecology in the British flora have heavily influenced plant ecophysiology, but many of his and subsequent authors' analyses have significant limitations because of inadequate statistical methods. For example, few analyses until recently account for phylogenetic nonindependence \citep{Felsenstein_1985}, which can strongly influence inferences on stomatal traits and growth form \citep{Kelly_Beerling_1995}. A species-level phylogeny of the entire British flora \citep{Lim_etal_2014} now allows me to rigorously analyze evolutionary relationships between stomatal ratio, light, and growth form.

>>>>

\begin{Schunk}
\begin{Soutput}
[1] TRUE
\end{Soutput}
\end{Schunk}

@@

\begin{itemize}
	\item In most plants, stomata are located only on the abaxial leaf surface (hypostomy), but many plants have stomata on both surfaces (amphistomy). Variation in stomatal ratio (the ratio of ab- and adaxial stomatal densities) is probably adaptive, but the ecological conditions that favor amphistomy are not well understood. In particular, high light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together.
	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, Raunki\ae r lifeform, and phylogenetic relationships for 372 species of British angiosperms. I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio.
	\item (return to this) key results: \el, growth form, and interaction are important
	\item I show for the first time that light and growth form interact to shape variation in stomatal ratio; amphistomy is advantageous in high light, but mostly for herbs. These results improve our understanding of the adaptive significance of stomatal ratio, use stomatal ratio as proxy for paleo vegetation, and as a target for crop improvement.
\end{itemize}

@@

To address these questions, I reanalyzed existing data on stomatal ratio, light habitat, and growth form in British angiosperms \citep{Salisbury_1927, Fitter_Peat_1994a, BEF} using phylogenetic comparative methods. The British angiosperm flora is well suited for these questions because this flora has been comprehensively surveyed for many ecologically important traits, meaning it is probably the least biased survey of stomatal trait variation. Salisbury's observations on stomata and ecology in the British flora have heavily influenced plant ecophysiology, but many of his and subsequent authors' analyses have significant limitations because of inadequate statistical methods. For example, few analyses until recently account for phylogenetic nonindependence \citep{Felsenstein_1985}, which can strongly influence inferences on stomatal traits and growth form \citep[this study did not consider light]{Kelly_Beerling_1995}. A species-level phylogeny of the entire British flora \citep{Lim_etal_2014} now allows me for the first time to rigorously analyze evolutionary relationships between stomatal ratio, light, and growth form.

% NOTES , for example, two herbaceous species adapted to different light conditions may not really be comparable since light has a strong impact on life history selection \citep{Franco_Silvertown_1996, Galloway_Etterson_2007}.
